---
title: "DallasCrimes"
author: "Shriram Rajasekar"
date: "2023-03-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chunk 1
```{r}
library(ggplot2)
library(stringr)
library(lubridate)
library(tidyverse)
library(dplyr)
dallas_crimes <- read.csv(file = "crime.csv", header = T)
phoenix_crimes <- read.csv(file = "arrestdetail.csv")
# cleaning the data

index <- which(dallas_crimes$Arrest.Year >= 2018 & dallas_crimes$Arrest.Year <= 2022)
dallas_crimes <- dallas_crimes[dallas_crimes$Arrest.Year >= 2018 & dallas_crimes$Arrest.Year <= 2022, ]
dallas_crimes <- arrange(dallas_crimes, Arrest.Year, Arrest.Date)

dallas_crimes$Arrest.Date <- as.Date(dallas_crimes$Arrest.Date, format = "%m/%d/%y", tryFormats = c("%m/%d/%y"))

phoenix_crimes$DATE_OCCUR <- as.Date(phoenix_crimes$DATE_OCCUR, format = "%m/%d/%Y")

phoenix_crimes <- dplyr::select(phoenix_crimes, -QTR, -UNIQUE_NAME_ID, -ARST_OFFICER,
                         -PRECINCT_NUM, -BEAT_NUM, -MAPGRID, -COUNCIL_DISTRICT)
phoenix_crimes <- filter(phoenix_crimes, YEAR < 2023)
phoenix_crimes <- na.omit(phoenix_crimes)
phoenix_crimes <- phoenix_crimes[seq(1, nrow(phoenix_crimes),4), ]
colnames(dallas_crimes) <- c("incidentnum", "year", "date", "time", 
                            "address", "zipcode", "latitude", "longitude", "city", 
                            "state", "dotw", "location", "weapon", "age",  "race", "sex", "related", "type")
colnames(phoenix_crimes) <- c("incidentnum", "date", "dotw", "month", "year", 
                              "type", "sex", "race", "ethnicity", "racegrp", "age",
                              "agegrp", "officersex", "officerrace", "officergrp", 
                              "address", "precinct", "beat", "felonycharges", "misdemeanorcharges", "othercharges", "unkowncharges",  "violentcharges", "propertycharges", "drugcharges", "assaultofficercharges")
rep_str <- c("1-" = "","2-" = "", "3-" = "", "4-" = "", "5-" = "", "6-" = "", "7-" = "")
phoenix_crimes$dotw <- str_replace_all(phoenix_crimes$dotw, rep_str)
rep_str2 <- c("0-" = "", "1-" = "","2-" = "", "3-" = "", "4-" = "", "5-" = "", "6-" = "", "7-" = "", "8-" = "", "9-" = "", "10-" = "", "11-" = "", "12-" = "")
phoenix_crimes$month <- str_replace_all(phoenix_crimes$month, rep_str2)
phoenix_crimes$month <- str_replace(phoenix_crimes$month,"0", "")
phoenix_crimes$month <- str_replace(phoenix_crimes$month,"1", "")

dallas_crimes$month <- month(ymd(dallas_crimes$date))
dallas_crimes$month <- month.abb[dallas_crimes$month]
dallas_crimes<- dallas_crimes[order(dallas_crimes$date), ]
dallas_crimes <- dallas_crimes[-c(1,2), ]
capwords <- function(s, strict = FALSE) {
    cap <- function(s) paste(toupper(substring(s, 1, 1)),
                  {s <- substring(s, 2); if(strict) tolower(s) else s},
                             sep = "", collapse = " " )
    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
}
phoenix_crimes$month <- substr(phoenix_crimes$month, 1, 3)
phoenix_crimes$month <- capwords(tolower((phoenix_crimes$month)))
phoenix_crimes$dotw <- substr(phoenix_crimes$dotw, 1, 3)
phoenix_crimes$dotw <- capwords(tolower(phoenix_crimes$dotw))

```

Chunk 2
Compare the two crime numbers and show that the ratio of crimes is very similar given the population of the two cities. So essentially since the crime rates are similar, it is important to consider the types of crimes, the timing, and location when deciding to move to the city. Note that dallas crimes in 2022 was only recorded up to 2022-02-21
```{r}
# plotting number of crimes
ggplot(data = phoenix_crimes) +
  geom_bar(mapping = aes(x = year), fill = "darkgreen") +
  xlab("Years") +
  ylab("Crimes") +
  ggtitle("Pheonix Crimes")
ggplot(data = dallas_crimes) +
  geom_bar(mapping = aes(x = year), fill = "red") +
  xlab("Years") +
  ylab("Crimes") + 
  ggtitle("Dallas Crimes")
```


Chunk 3
```{r}
ggplot(data = dallas_crimes) +
  geom_bar(stat = "count", aes(x = fct_inorder(month)), fill = "red") +
  ggtitle("Crimes in Dallas per month") +
  labs(x = "Month", y = "Number of Crimes")


ggplot(data = phoenix_crimes) +
  geom_bar(stat = "count", aes(x = fct_inorder(month)), fill = "darkgreen") +
  ggtitle("Crimes in phoenix per month") +
  labs(x = "Month", y = "Number of Crimes")
```

Chunk 4
```{r}
dallas_barplot <- ggplot(dallas_crimes, aes(x = fct_inorder(dotw))) + 
  geom_bar(stat = "count", aes(fill = dotw)) +
  labs(x = "Day of Week", y = "Number of Crimes") + 
  ggtitle("Number of crimes in Dallas by day of the week")

dallas_barplot_dotw <- dallas_barplot + 
  scale_fill_manual(values = c("blue", "green", "red", "orange", "purple", "brown", "pink"))

dallas_barplot_dotw

phoenix_barplot <- ggplot(phoenix_crimes, aes(x = fct_inorder(dotw))) + 
  geom_bar(stat = "count", aes(fill = dotw)) +
  labs(x = "Day of Week", y = "Number of Crimes") + 
  ggtitle("Number of crimes in phoenix by day of the week")

phoenix_barplot_dotw <- phoenix_barplot + 
  scale_fill_manual(values = c("blue", "green", "red", "orange", "purple", "brown", "pink")) 

phoenix_barplot_dotw
```

Chunk 5
```{r}
ggplot(data = dallas_crimes) +
  geom_line(stat = "count", aes(x = date)) +
  ggtitle("Crimes trend overtime for Dallas") +
  labs(x = "Month", y = "Number of Crimes")


ggplot(data = phoenix_crimes) +
  geom_line(stat = "count", aes(x = date)) +
  ggtitle("Crime trend overtime for phoenix") +
  labs(x = "Month", y = "Number of Crimes")
```

Chunk 6
General crime trend... Dallas crime rate may be more sporadic, Phoenix may be more but less spikes. maybe we can add some line to indicate when christmas is or other major holidays
```{r}
ggplot(data = dallas_crimes) +
  geom_line(stat = "count", aes(x = date)) +
  ggtitle("Crimes trend overtime for Dallas") +
  labs(x = "Month", y = "Number of Crimes")


ggplot(data = phoenix_crimes) +
  geom_line(stat = "count", aes(x = date)) +
  ggtitle("Crime trend overtime for phoenix") +
  labs(x = "Month", y = "Number of Crimes")
```

Chunk 7
Drug related crimes/ ages committing the crimes
```{r}
ggplot(data = dallas_crimes) +
  geom_freqpoly(aes(x = age, color = drugrelated), bins = 20) +
  ggtitle("Criminals Age for Dallas") +
  labs(x = "Age", y = "Number of Crimes")
ggplot(data = phoenix_crimes) +
  geom_freqpoly(aes(x = age, color = drugcharges), bins = 20) +
  ggtitle("Criminals Age for phoenix") +
  labs(x = "Age", y = "Number of Crimes")
```



Chunk 8
```{r}
dallas_crimes_highway <- filter(dallas_crimes, grepl("Highway, Street, Alley ETC", location))
dallas_crimes_highway <- filter(dallas_crimes_highway, sex != "Unknown")

ggplot(dallas_crimes_highway, aes(x = age, y = location, fill = sex)) +
  geom_boxplot() +
  labs(x = "Age", y = "Location") +
  ggtitle("Boxplot of location and age")


pheonix_crimes <- pheonix_crimes[, which(!is.na(colnames(pheonix_crimes)) & colnames(pheonix_crimes) != "")]

phoenix_crimes_downtown <- filter(pheonix_crimes, grepl("Central City", precinct))


ggplot(phoenix_crimes_downtown, aes(x = age, y = precinct, fill = sex)) +
  geom_boxplot() +
  labs(x = "Age", y = "Location") +
  ggtitle("Boxplot of location and age")

```

Chunk 9 
Violent crimes in specific locations/zipcodes
```{r}
dallas_crimes <- filter(dallas_crimes, zipcode > 70000)
ggplot(data = dallas_crimes) +
  geom_jitter(aes(x = zipcode, y = weapon))
ggplot(data = phoenix_crimes) +
  geom_jitter(aes(x = precinct, y = violentcharges))
```

```{r}
# Create a box plot of race vs felony charges
ggplot(data = phoenix_crimes, aes(x = racegrp, y = felonycharges)) +
  geom_boxplot(fill = "steelblue", alpha = 0.8) +
  labs(title = "Boxplot of Race vs Felony Charges in Phoenix", x = "Race", y = "Felony Charges") +
  theme_minimal()


```











